Supermartingales and Termination Analysis of Statistical PCF

### 2020-12-28 LICS21/confluent.tex
Mainly LaTeX reformatting to fit 2-column style

\section{Confluent trace semantics}
\label{sec:confluent}

@@

A \emph{position} 
%\akr{I'm not really sure how to phrase the citation, but this is adapted from \cite{Kennaway96infinitarylambda}.} 
%\lo{The idea is common enough. I think no citation necessary.}
is a finite sequence of steps into a term, defined inductively as
\begin{align*}
\alpha ::= \cdot \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha \mid \tscore ; \alpha.
\end{align*}
The \emph{subterm of $M$ at a position $\alpha$}, denoted $M \mid \alpha$, is defined as
\begin{align*}
M \mid \cdot & = M \\
\lambda x. M \mid \lambda ; \alpha & = M \mid \alpha \\
M_1 M_2 \mid @_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2 \\
\underline f(M_1,\dots,M_n) \mid \underline f_i ; \alpha & = M_i \mid \alpha \quad \text{for }i \leq n \\
\tY M \mid \tY ; \alpha & = M \mid \alpha \\
\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2,3 \\
\tscore(M) \mid \tscore ; \alpha & = M \mid \alpha
\end{align*}
so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda$ is undefined). 
A position such that $M\mid \alpha$ does exist is said to \emph{occur} in $M$. 
%\changed[lo]{
\emph{Substitution} of $N$ at position $\alpha$ in $M$, written $M[N/\alpha]$, is defined similarly.
For example, let 
\[
M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})
\qquad
\alpha =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
\]
then 
\(
M[\tsample / \alpha] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
\)
%}

@@

With this notation, a more general reduction relation $\to$ can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
%\lo{On first reading, I struggled to parse the following clauses.}
\begin{definition}
\label{def:more general red}
The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $\alpha$ in the term $M$:
\begin{align*}
  \text{if } M \mid \alpha = (\lambda x.N) V,\ & M \to M[N[V/x]/\alpha] \\
  \text{if } M \mid \alpha = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/\alpha] \\
  \text{if } M \mid \alpha = \tY \lambda x. N,\ & M \to M[\lambda z. N[(\tY \lambda x. N)/x] z/\alpha] \text{ where $z$ is not free in $N$}\\
  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_1/\alpha] \text{ where }r < 0 \\
  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_2/\alpha] \text{ where }r \geq 0 \\
  \text{if } M \mid \alpha = \tsample & \text{ and $\lambda$ does not occur after $@_2$ or $\tY$ in $\alpha$},\\ & M \to M[\underline r/\alpha] \text{ where } r \in [0,1] \\
  \text{if } M \mid \alpha = \tscore(\underline r),\ & M \to M[\underline r/\alpha].
\end{align*}
In each of these cases, $M \mid \alpha$ is the \emph{redex}, and reduction takes place at $\alpha$.
\end{definition}

@@

  The effect of this rearrangement is (ignoring the $\sim_c$s) to swap the order of the $\sim_\uparrow$ and the $\sim_\downarrow$, except that if the positions of one of these reductions is inside the other, that inner $\sim_p$ may be duplicated. If the positions are disjoint, both of them are unchanged (corresponding to case a of $\sim_c$), and if one is inside the other, then the outer position is unchanged and the other resultant positions, although they aren't equal to the original inner position, are still inside the outer position. Taking the sub-sequence of one $\sim_\uparrow$ followed by a parallel sequence of $\sim_\downarrow$s, this rearrangement can be repeatedly applied to move the $\sim_\uparrow$ further along the sequence. If at some point a $\sim_\uparrow$ and $\sim_\downarrow$ match (by relating the same (skeleton, position) pairs in opposite directions), they may be removed and the process may stop early. The $\sim_\uparrow$ may be duplicated if it passes a $\sim_\downarrow$ whose position is outside of the $\sim_\uparrow$'s position, but by the assumption that the sequence of $\sim_\downarrow$s is initially parallel, no position inside of this occurs later in the sequence, therefore all of the resultant $\sim_\uparrow$s pass the remaining $\sim_\downarrow$s without changing or duplicating them. The $\sim_\uparrow$s may be further duplicated, but the number of steps left in this process is bounded by the product, for all of the remaining $\sim_\downarrow$s, of $2 + \text{the number of occurrences of the variable relevant to the reduction}$, because for each $\sim_\downarrow$, that is more than the number of duplicates it could produce for any $\sim_\uparrow$ that passes it. It is also possible (earlier) for some of the $\sim_\downarrow$s to be duplicated, but the only case where this process would not terminate, that both the $\sim_\downarrow$s and the $\sim_\uparrow$s continue being duplicated forever so that the number of switches left to do never decreases, is prevented by the parallelness condition, as all of the duplications of $\sim_\downarrow$s occur before all of the duplications of $\sim_\uparrow$s.

@@

Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
\changed[lo]{We write an element of $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ as $(M', s)$ where the \emph{term} $M' \in \Lambda^0$ and $s \in  I^{L_s(M')}$.}
\begin{align*}
& (M,s) \Rightarrow (N,s \circ i(M \to N)) \text{ if either} \\
& \qquad \text{$M \to N$ and the redex is not $\tsample$, or} \\
& \qquad \text{$M \mid \alpha = \tsample$, $N = M[\underline{s(Sk(M),\alpha)}/\alpha]$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $\alpha$}
\end{align*}

@@

R_s & = \{\text{the skeleton of } O \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) = \tsample, O \to N[r]\} \\
R_d & = \{\text{the skeleton of } O \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) \neq \tsample, O \to N[r]\}.

@@

@@

= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in S :\\
    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\})\\
    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in S :\\
    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \\

@@

= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} :\\
    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S)\\
    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} :\\
    & (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S) \\

@@

The domain of definition of the ranking functions also needs to be changed from the reachable terms $Rch(M)$ to the \emph{reachable terms with respect to the reduction strategy $r$},
\[
Rch_r(M) = \{N \mid \exists n, N_0, \dots, N_n: N_0 = M, N_n = N, N_i \to N_{i+1} \text{ at } r(N_i)\}.
\]

>>>>

% !TEX root = main.tex
\section{Confluent trace semantics}
\label{sec:confluent}

@@

A \emph{position} is a finite sequence of steps into a term, defined inductively as
\begin{align*}
\alpha ::= \cdot \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \\
\mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha \mid \tscore ; \alpha.
\end{align*}
The \emph{subterm of $M$ at a position $\alpha$}, denoted $M \mid \alpha$, is defined as
\begin{align*}
M \mid \cdot & = M \\
\lambda x. M \mid \lambda ; \alpha & = M \mid \alpha \\
M_1 M_2 \mid @_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2 \\
\underline f(M_1,\dots,M_n) \mid \underline f_i ; \alpha & = M_i \mid \alpha \quad \text{for }i \leq n \\
\tY M \mid \tY ; \alpha & = M \mid \alpha \\
\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2,3 \\
\tscore(M) \mid \tscore ; \alpha & = M \mid \alpha
\end{align*}
so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda$ is undefined). 
A position such that $M\mid \alpha$ does exist is said to \emph{occur} in $M$. 
%\changed[lo]{
\emph{Substitution} of $N$ at position $\alpha$ in $M$, written $M[N/\alpha]$, is defined similarly.
For example, let 
\[
M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})
\qquad
\alpha =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
\]
then 
\(
M[\tsample / \alpha] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
\)
%}

@@

With this notation, a more general reduction relation $\to$ can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
%\lo{On first reading, I struggled to parse the following clauses.}
\begin{definition}
\label{def:more general red}
The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $\alpha$ in the term $M$:
\begin{align*}
  \text{if } M \mid \alpha = (\lambda x.N) V,\ & M \to M[N[V/x]/\alpha] \\
  \text{if } M \mid \alpha = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/\alpha] \\
  \text{if } M \mid \alpha = \tY \lambda x. N,\ & M \to M[\lambda z. N[(\tY \lambda x. N)/x] z/\alpha]\\ \text{where }z&\text{ is not free in $N$}\\
  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_1/\alpha] \text{ where }r < 0 \\
  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_2/\alpha] \text{ where }r \geq 0 \\
  \text{if } M \mid \alpha = \tsample \text{ and $\lambda$ does}&\text{ not occur after $@_2$ or $\tY$ in $\alpha$},\\ & M \to M[\underline r/\alpha] \text{ where } r \in [0,1] \\
  \text{if } M \mid \alpha = \tscore(\underline r),\ & M \to M[\underline r/\alpha].
\end{align*}
In each of these cases, $M \mid \alpha$ is the \emph{redex}, and reduction takes place at $\alpha$.
\end{definition}

@@

  The effect of this rearrangement is (ignoring the $\sim_c$s) to swap the order of the $\sim_\uparrow$ and the $\sim_\downarrow$, except that if the positions of one of these reductions is inside the other, that inner $\sim_p$ may be duplicated. If the positions are disjoint, both of them are unchanged (corresponding to case a of $\sim_c$), and if one is inside the other, then the outer position is unchanged and the other resultant positions, although they aren't equal to the original inner position, are still inside the outer position. Taking the sub-sequence of one $\sim_\uparrow$ followed by a parallel sequence of $\sim_\downarrow$s, this rearrangement can be repeatedly applied to move the $\sim_\uparrow$ further along the sequence. If at some point a $\sim_\uparrow$ and $\sim_\downarrow$ match (by relating the same (skeleton, position) pairs in opposite directions), they may be removed and the process may stop early. The $\sim_\uparrow$ may be duplicated if it passes a $\sim_\downarrow$ whose position is outside of the $\sim_\uparrow$'s position, but by the assumption that the sequence of $\sim_\downarrow$s is initially parallel, no position inside of this occurs later in the sequence, therefore all of the resultant $\sim_\uparrow$s pass the remaining $\sim_\downarrow$s without changing or duplicating them. The $\sim_\uparrow$s may be further duplicated, but the number of steps left in this process is bounded by the product, for all of the remaining $\sim_\downarrow$s, of $2 +$ the number of occurrences of the variable relevant to the reduction, because for each $\sim_\downarrow$, that is more than the number of duplicates it could produce for any $\sim_\uparrow$ that passes it. It is also possible (earlier) for some of the $\sim_\downarrow$s to be duplicated, but the only case where this process would not terminate, that both the $\sim_\downarrow$s and the $\sim_\uparrow$s continue being duplicated forever so that the number of switches left to do never decreases, is prevented by the parallelness condition, as all of the duplications of $\sim_\downarrow$s occur before all of the duplications of $\sim_\uparrow$s.

@@

Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
\changed[lo]{We write an element of $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ as $(M', s)$ where the \emph{term} $M' \in \Lambda^0$ and $s \in  I^{L_s(M')}$.}
\begin{align*}
& (M,s) \Rightarrow (N,s \circ i(M \to N)) \text{ if $M \to N$ at $\alpha$ and either} \\
& \qquad \text{the redex is not $\tsample$, or} \\
& \qquad \text{$M \mid \alpha = \tsample$ and $N = M[\underline{s(Sk(M),\alpha)}/\alpha]$}
\end{align*}

@@

R_s & = \{Sk(O) \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) = \tsample, O \to N[r]\} \\
R_d & = \{Sk(O) \mid O \in Rch(M), r \in \mathbb R^k, O | f(O) \neq \tsample, O \to N[r]\}.

@@

\clearpage

@@

= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in S : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\})\\
    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in S : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \\

@@

= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S)\\
    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S) \\

@@

The domain of definition of the ranking functions also needs to be changed from the reachable terms $Rch(M)$ to the \emph{reachable terms with respect to the reduction strategy $r$},
\begin{align*}
& Rch_r(M) = \\ & \{N \mid \exists n, (N_i)_{0 \leq i \leq n}: N_0 = M, N_n = N, N_i \to N_{i+1} \text{ at } r(N_i)\}.
\end{align*}
